---
title: Meteorite Landings
author: Santiago Gualchi
format:
  html:
    df-print: kable
    fig-width: 6
    fig-height: 6
    toc: true
    toc-depth: 3
execute: 
  warning: false
---

## Introduction

## Packages

```{r}
library(dplyr)
library(forcats)
library(ggimage)
library(ggplot2)
library(ggtext)
library(glue)
library(readr)
library(stringr)
library(tibble)
library(tidyr)

source("R/theming.R")
```

## Data

We'll use the web-scraped data if available, or the alternative data otherwise
(see README.md).

```{r}
file_path <- if (file.exists("nogit/tidy-data.csv")) {
  "nogit/tidy-data.csv"
} else {
  "data/meteorite-landings/meteorite-landings.csv"
}

data_raw <- read_csv(
  file_path,
  na = "",
  col_types = cols_only(
    name = col_character(),
    nametype = col_factor(),
    # recclass = col_factor(),
    mass = col_double(),
    fall = col_factor(),
    year = col_integer(),
    reclat = col_double(),
    reclong = col_double(),
    place = col_character()
  )
) |> 
  # This mutates are performed to transform the alternative data into a format
  # compatible with the web-scraped data:
  mutate(
    nametype = fct_recode(nametype, Official = "Valid"),
    fall = fct_recode(fall, Fall = "Fell", Find = "Found")
  ) |> 
  # This left_join is performed in order to ensure that there's a place column
  # in the alternative data.
  left_join(
    tibble(name = "INEXISTENT-NAME-SANTIGUALCHI.DATA", place = ""),
    by = "name",
    # We include the suffixes because, for the web-scraped data, there'll be
    # two place columns. This way, we preserve the original place column name
    # and distinguish the dummy column with the suffix "_drop".
    suffix = c("", "_drop")
  ) |> 
  # Now we can drop the dummy column if present.
  select(-ends_with("_drop"))

head(data_raw, 10)
```

Now we'll do some tidying:

```{r}
data_tidy <- data_raw |> 
  # Rename to have more descriptive names.
  rename(
    # class = recclass,
    latitude = reclat,
    longitude = reclong
  ) |> 
  # Remove data-points of on-going year.
  filter(year < max(year, na.rm = TRUE)) |> 
  # Keep only meteorites (other nametypes are not meteorites).
  filter(nametype %in% c("Official", "Provisional")) |> 
  mutate(
    # Remove meaningless trailing characters from names.
    name = str_replace(name, r"{\s*\*\*$}", ""),
    # Remove meteorites on null island.
    latitude = if_else(latitude == 0 & longitude == 0, NA_real_, latitude),
    longitude = if_else(latitude == 0 & longitude == 0, NA_real_, longitude)
  )

head(data_tidy, 10)
```

Get counts by year:

```{r}
data_counts_by_year <- data_tidy |> 
  group_by(year) |> 
  summarize(n = n())

head(data_counts_by_year, 10)
```


## Cover

This will be the cover of the instagram post. I'm using `{ggplot2}` to create
it, to have a consistent format across all slides:

```{r}
ggplot(tibble(0)) +
  geom_textbox(
    label = "Meteorites:<br>Landings<br>And Discoveries",
    x = 0,
    y = 0,
    width = unit(5, "inch"),
    size = 14,
    fontface = "bold",
    color = NA,
    text.color = color_bg,
    fill = NA,
    lineheight = 1.3,
    halign = 1,
    hjust = .5,
    vjust = .5
  ) + 
  scale_x_continuous(limits = c(-9, 9)) +
  scale_y_continuous(limits = c(-9, 9)) +
  theme_cover
```


## 200 Years Of Meteoritics

Reports of falling meteorites and discoveries on land extend thousands of years
into the past, but the mysteries of their origin and nature were not always
understood. Over time, various explanations have been proposed, ranging from
religious and magical to more scientific theories. Some cultures revered
meteorites as divine gifts, while others attributed their existence to volcanic
eruptions or atmospheric phenomena. Many people even dismissed the existence
of meteorites altogether, discrediting any mention of them as mere folklore.

Instead, the idea that meteorites come from outer space was disregarded by
the scientific community for many years, as it was believed that no small
objects existed in interplanetary space. However, in 1794 Ernst Chladni
suggested that the physical properties of these objects could only be accounted
for by their extraterrestrial origin. Chladni's idea amounted to scientific
heresy at the time, but subsequent research, along with some conveniently-timed
meteorite sightings, ultimately led to widespread recognition that
meteorites are indeed extraterrestrial in 1803.

Since then, interest in space greatly increased and numerous technological
breakthroughs were made. In 1974,
`r data_counts_by_year[data_counts_by_year$year == 1974, ]$n` meteorites were
discovered, a dramatic increase from the previous record of
`r max(data_counts_by_year[data_counts_by_year$year < 1974, ]$n)` in
`r max(data_counts_by_year[data_counts_by_year$year < 1974, ]$year)`. But the
current record was reached in the year
`r data_counts_by_year[max(data_counts_by_year$n) == data_counts_by_year$n, ]$year`,
with 
`r data_counts_by_year[max(data_counts_by_year$n) == data_counts_by_year$n, ]$n`
meteorites discovered. As of the beginning of 2023, over `r nrow(data_tidy)`
meteorites are known.


```{r}
data_counts_by_year_since_1800 <- data_counts_by_year |> 
  filter(year >= 1800)

milestones <- tribble(
  ~year, ~milestone,
  1803,  "Recognition of meteorites<br>as extraterrestrial",
  1974,  "",
  1979,  "",
  2000,  "Current record"
)

ggplot(
  data_counts_by_year_since_1800,
  mapping = aes(
    x = year,
    y = n
  )
) +
  geom_vline(
    data = milestones,
    mapping = aes(
      xintercept = year
    ),
    linetype = "longdash",
    color = color_grey_1
  ) +
  geom_richtext(
    data = milestones,
    mapping = aes(
      x = year + c(3, -3, 3, 3),
      label = paste0(
        "<span style='font-size:12pt;'>**", year, "**</span><br>",
        "<span style='font-size:11pt;'>", milestone, "</span><br>"
      )
    ),
    y = max(data_counts_by_year_since_1800$n) + 1800,
    color = color_grey_3,
    fill = "transparent",
    label.color = "transparent",
    hjust = c(0, 1, 0, 0),
    vjust = 1,
    label.padding = unit(0, "pt")
  ) +
  geom_col(
    fill = color_accent_light,
    # This overlaps, but I want the bars to be more visible.
    width = 2
  ) +
  scale_x_continuous(
    limits = c(1800, 2023),
    breaks = seq(1800, 2023, by = 50) |> c(2023) |> unique(),
    expand = expansion(mult = c(.08, .08))
  ) +
  scale_y_continuous(
    breaks = NULL,
    expand = expansion(mult = 0, add = c(0, 2000))
  ) +
  labs(
    title = "200 Hundred Years Of Meteoritics",
    subtitle = "Most discoveries happened after the Space Age began",
    x = NULL,
    y = NULL,
    caption = "Source: The Meteoritical Bulletin Database"
  ) +
  theme_default +
  theme_x_line
```

## Over X Found In 10 Years

DESCRIPTION PENDING.

```{r}
data_counts_by_year_last_10_years <- data_counts_by_year |>
  filter(year > (max(year) - 10))

# This function formats x. It truncates x to only keep the number signif of
# significative digits (even for whole numbers). For example,
# pretty_number(12345) returns 12000. Further arguments are passed to prettyNum.
pretty_number <- function(x, signif = 2, ...) {
  (x - x %% 10 ^ (floor(log10(x)) - (signif - 1))) |> 
    prettyNum(...)
}

ggplot(
  data_counts_by_year_last_10_years,
  mapping = aes(
    x = year,
    y = n
  )
) +
  geom_col(
    fill = color_accent_light,
    width = .7
  ) +
  geom_text(
    aes(
      label = n
    ),
    size = 4,
    fontface = "bold",
    color = color_bg,
    nudge_y = -60
  ) +
  scale_x_continuous(
    breaks = data_counts_by_year_last_10_years$year,
    expand = expansion(mult = 0)
  ) +
  scale_y_continuous(
    breaks = NULL,
    expand = expansion(mult = 0)
  ) +
  labs(
    title = glue(
      "Over ",
      pretty_number(sum(data_counts_by_year_last_10_years$n), big.mark = ","),
      " Found In 10 Years", 
    ),
    subtitle = glue(
      "And many are still in the process of being recognized"
    ),
    x = NULL,
    y = NULL,
    caption = "Source: The Meteoritical Bulletin Database"
  ) +
  theme_default
```

## A Global Phenomenon

DESCRIPTION PENDING

```{r}
data_positions <- data_tidy |> 
  filter(longitude >= -180 & longitude <= 180) |> 
  replace_na(
    list(
      mass = median(data_tidy$mass, na.rm = TRUE)
    )
  ) |> 
  arrange(desc(mass))
```

```{r}

get_meteorite_textbox <- function(data, nrow, x, y) {
  geom_textbox(
    data = data[nrow, ],
    mapping = aes(
      label = glue(
        "<span style='font-size:12pt;'>**{name}** ({year})</span><br>",
        "<span style='font-size:11pt;'>", place, "<br>",
        "{prettyNum(mass / 1e3, big.mark = ',')} kg</span>"
      )
    ),
    x = x,
    y = y,
    box.r = unit(0, "pt"),
    fill = color_bg,
    box.color = color_accent_dark,
    box.size = .5,
    color = color_grey_4
  )
}

world <- map_data("world")

ggplot(data_positions) +
  geom_map(
    data = world,
    map = world,
    mapping = aes(
      map_id = region
    ),
    color = color_grey_1,
    fill = color_grey_1
  ) +
  geom_point(
    aes(
      x = longitude,
      y = latitude,
      size = mass
    ),
    color = color_grey_2,
    shape = 1,
    alpha = 1/2
  ) +
  geom_segment(
    data = data_positions[1:4, ],
    mapping = aes(
      x = longitude,
      y = latitude,
      xend = longitude
    ),
    yend = c(135, -135, 135, -135),
    color = color_accent_dark
  ) +
  geom_segment(
    data = data_positions[1:4, ],
    mapping = aes(
      x = longitude
    ),
    y = c(135, -135, 135, -135),
    xend = c(110, 110, -110, -110),
    yend = c(135, -135, 135, -135),
    color = color_accent_dark
  ) +
  get_meteorite_textbox(
    data_positions,
    nrow = 1,
    x = 110,
    y = 135
  ) +
  get_meteorite_textbox(
    data_positions,
    nrow = 2,
    x = 110,
    y = -135
  ) +
  get_meteorite_textbox(
    data_positions,
    nrow = 3,
    x = -110,
    y = 135
  ) +
  get_meteorite_textbox(
    data_positions,
    nrow = 4,
    x = -110,
    y = -135
  ) +
  geom_point(
    data = data_positions[1:4, ],
    aes(
      x = longitude,
      y = latitude,
      size = mass
    ),
    color = color_accent_dark,
    fill = color_accent_light,
    shape = 21,
    stroke = .65
  ) +
  guides(
    size = "none"
  ) +
  scale_x_continuous(
    breaks = NULL,
    expand = expansion(mult = 0),
    limits = c(-180, 180)
  ) +
  scale_y_continuous(
    breaks = NULL,
    expand = expansion(mult = 0),
    # latitude is bound -90 to 90, but we're including extra space for the
    # textboxes.
    limits = c(-180, 180)
  ) +
  labs(
    title = "A Global Phenomenon",
    subtitle =  "Meteorites have been striking Earth all over the globe",
    x = NULL,
    y = NULL,
    caption = "Source: The Meteoritical Bulletin Database"
  ) +
  theme_default
```

## What's in a meteorite?

```{r}
# hoba_fun_facts <- data.frame(
#   fun_fact = c(
#     "Largest known meteorite found intact. It weighs 60 tons.",
#     "Iron meteorite. Composed of 84% iron and 16% nickel.",
#     "Landed on Namibia about 80,000 years ago."
#   )
# )
# 
# ggplot(hoba_composition) +
#   annotation_raster(
#     img,
#     xmin = -Inf,
#     xmax = Inf,
#     ymin = -Inf,
#     ymax = Inf
#   )
#   # geom_image(
#   #   image = "resources/hoba.svg",
#   #   x = 0,
#   #   y = 0,
#   #   size = 1
#   # ) +
#   scale_fill_discrete(type = c(color_grey_2, color_grey_2)) +
#   xlim(c(0.2, 8 + .5)) +
#   guides(fill = "none") +
#   labs(
#     title = "A Closer Look",
#     subtitle = "Characteristics of the Hoba meteorite"
#   ) +
#   theme_default +
#   theme(
#     axis.title = element_blank(),
#     axis.text = element_blank(),
#     axis.text.x = element_blank()
#   )
```